---
title: "Geology Analyte Analysis App"
author: "Ford Fishman"
date: "6/28/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

This document was written with the purpose of explaining the code used in the following app in relatively simple terms:<https://fordfishman.shinyapps.io/IPC-analysis-app/>.

## Setup

First, we attach the necessary packages to read and export excel files and to manipulate them in R. Make sure to install each of these before running the standalone script.

```{r}
suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(library(argparse))
suppressPackageStartupMessages(library(readxl))
suppressPackageStartupMessages(library(xlsx))
```

This next segment is the computational stand-in for the user picking files and changing default settings in the actual app. It's in the standalone script, but it's not a part of the Shiny app.

```{r}
path = "~/Documents/Geology_program/" 
file <- "/Users/fordfishman/Documents/Geology_program/Field Foams 4,12,13,14 rerun,145,147,156,157,158,160,161.xlsx"
limits <- "/Users/fordfishman/Documents/Geology_program/MDL-LOQ limits.xlsx"
weights <- "/Users/fordfishman/Documents/Geology_program/Masses of dried field foams (mod 1-06.27.19).xlsx"
rsquare_cutoff <- 0.94
sample_start <- "A"
sample_end <- "C"
output <- "/Users/fordfishman/Documents/Geology_program/IPC_analysis.xlsx"
load("/Users/fordfishman/Documents/Geology_program/IPC-analysis-app/data/IPC_data.Rdata") # This data includes background foam concentrations and weights
```

The following code imports the main data sheet ("Conc. in Calib Units") and makes sure it is formatted correctly. `gsub()` searches a given set of characters and replaces specified characters with other specified characters.

```{r}
# Saves the sheet with concentration data into a dataframe
field_foams_concentrations_df <- read_excel(path = file, sheet = "Conc. in Calib Units")
# Convert sample number from a character string to an integer
field_foams_concentrations_df$Sample <- as.integer(field_foams_concentrations_df$..1)
# Rename some columns to remove ambiguous characters and names
colnames(field_foams_concentrations_df) <- gsub("\r", "", x = colnames(field_foams_concentrations_df))
colnames(field_foams_concentrations_df) <- gsub("\n", "", x = colnames(field_foams_concentrations_df))
colnames(field_foams_concentrations_df) <- gsub(" ", "", x = colnames(field_foams_concentrations_df))
colnames(field_foams_concentrations_df) <- gsub("mg", "µg", x = colnames(field_foams_concentrations_df))
colnames(field_foams_concentrations_df) <- gsub("As193\\.69\\(", "As193\\.696\\(", x = colnames(field_foams_concentrations_df))
```

## Drift Correction

Analyte values drift as the run continues. Therefore, correcting this drift is the first matter of business. 

The internal standards sheet is first extracted and edited. This sheet contains sample numbers and corresponding yttrium concentrations (`Y_levels`). 

The internal standards levels used for drift correction only correspond to the IPC-500 samples. `important_Y` contains this subset of Y concentrations.

```{r}
# Saves sheet with Y standard information
internal_standards_df <- read_excel(path = file, sheet = "Internal Standards")

# Adds Y information to concentration dataframe
colnames(internal_standards_df) <- c("RowIndex", "Y_levels")
field_foams_concentrations_df <- mutate(field_foams_concentrations_df, Y_levels = internal_standards_df$Y_levels)

# Alter sample names to remove " " to remove inconsistent naming.
field_foams_concentrations_df$SampleId <- gsub(" ", "", field_foams_concentrations_df$SampleId)

# Grab Y values for IPC-500 runs
important_Y <- subset(field_foams_concentrations_df, 
                      startsWith(SampleId, "IPC-500"), 
                      select = c(Sample, Y_levels))
```

The drift correction is done by making one or multiple linear regressions. Using the important Y values (dependent variable) and the sample number (independent variable), the program creates a linear model. The slope of this model represents the drift of the concentrations. 

As the machine's drift behavior can change throughout a run, one single model may poorly represent the drift behavior. Using the r-squared listed above as a minimum r-squared cutoff value for a given model, the program will subset from the important_Y dataset until it can make a model with an r-square at or exceeding the minimum. Y values at the end of the run will be removed first. The minimum r-squared value can be set at 1, requiring a perfect fit. 

For data points not used, this same approach is repeated until a model is found with an appropriate r-squared. This means that any given run can have anywhere from 1 to $n$-1 models representing the drift behavior, where $n$ is the number of IPC-500 samples. 

As $n$ is not constant from run to run, the max number of regressions necessary in a given run is also variable. The next segment of code simply makes names for variables representing the group of important Y values and their corresponding linear models (group1, lm1; group2, lm2; ...; group(n-1), lm(n-1)).

```{r}
# The following is just making variable names for the drift correction regressions
reg_group <- c() # Will contain names of subsets of data
lm_name <- c() # Will contains names of regressions
for (i in 1:(nrow(important_Y)-1)){
  group_name <- paste("group", as.character(i), sep = "")
  model <- paste("lm", as.character(i), sep = "")
  lm_name <- c(lm_name, model)
  reg_group <- c(reg_group, group_name)
}
```

The following does the process described above takes place in the code below. The `get()` function is used to call objects that can have different names. For instance, the current slope can be `m1`, `m2`, ..., `m(n-1)` depending on which regression it corresponds with.

The names of the equations of the regressions are stored in `functions_list` with the corresponding sample numbers that they cover (`y1` covers 1-20, `y2` covers 21-45). 

```{r}
# Getting linear models for the drift corrections
functions_list <- list() # A list of model names and the sample numbers they correspond with
START <-  1
END <- nrow(important_Y) 

for (i in 1:length(lm_name)){
  # Using all data for the regression at first
  assign(reg_group[i], important_Y[START:END,])
  assign(lm_name[i], lm(data = get(reg_group[i]), formula = Y_levels ~ `Sample`))
  # Removes the last entry in the set of data until the r-squared is above the cutoff
  # Can end up only having 2 data points, where r-squared is 1
  below_sig <- summary(get(lm_name[i]))$r.squared < rsquare_cutoff
  # below_sig will be NA if there is perfect fit
  if (is.na(below_sig)) {
    below_sig <- F
  }
  # While the r-squared value is below the threshold, 
  # pop a Y value from the end of the current group, do another regression
  while (below_sig) {
    END <- END - 1
    assign(reg_group[i], important_Y[START:END,])
    assign(lm_name[i], lm(data = get(reg_group[i]), formula = Y_levels ~ `Sample`))
    below_sig <- summary(get(lm_name[i]))$r.squared < rsquare_cutoff
    if (is.na(below_sig)) {
      below_sig <- F
    }
  }
  # Creates a linear formula when r-squared meets cutoff
  assign(paste("b", i, sep = ""), coef(get(lm_name[i]))[1])
  assign(paste("m", i, sep = ""), coef(get(lm_name[i]))[2])
  assign(paste("y", i, sep = ""), 
         function(x) 
           get(paste("m", i, sep = "")) * x + get(paste("b", i, sep = "")))
  ##Case 1: linear model covers all data
  if (START == 1 && END == nrow(important_Y)){
    functions_list[[paste("y", i, sep = "")]] <- c(1:nrow(field_foams_concentrations_df))
    ##Case 2: linear model starts at beginning of data but stops before end
  } else if (START == 1 && END != nrow(important_Y)) {
    functions_list[[paste("y", i, sep = "")]] <- c(1:as.integer(important_Y[END, 1]))
    ##Case 3: linear model doesn't start at beginning and it stops before end
  } else if (START != 1 && END != nrow(important_Y)){
    functions_list[[paste("y", i, sep = "")]] <- c((as.integer(important_Y[START, 1]) + 1):as.integer(important_Y[END, 1]))
    ##Case 4: linear model doesn't start at beginning but stops at end
  } else {
    functions_list[[paste("y", i, sep = "")]] <- c((as.integer(important_Y[START, 1]) + 1):nrow(field_foams_concentrations_df))
  }
  # The loop will start again where the last model left off
  # If the last model made it to the end, the loop will end
  START <- END
  END <- nrow(important_Y)
  if (START == END){
    break
  }
}
```

Now that we have equations modeling the drift behavior, we can predict the level of drift for each sample as a multiplier, and store each multiplier as a vector in `drift_multiplier`.

```{r}
drift_multiplier <- c()

# Iterates through and gives the correct drift multiplier to the correct sample
for (i in 1:length(functions_list)) {
  for (num in functions_list[i]){
    drift_multiplier <- c(drift_multiplier, get(names(functions_list[i]))(num))
  }
}
```

The drift correction is made by multiplying the concentrations by the inverse of the drift multiplier found above that is unique for each sample. The drift corrected data is stored in the dataframe `drift_corrections`.

```{r}
# Reorganization of dataframe
field_foams_concentrations_df <- mutate(field_foams_concentrations_df, drift_multiplier = drift_multiplier)
drift_corrections <-  data.frame(n = field_foams_concentrations_df$..1, id = field_foams_concentrations_df$SampleId)
for (i in names(field_foams_concentrations_df)) {
  if (is.numeric(field_foams_concentrations_df[[i]]) & endsWith(i, ")")) {
    drift_corrections[[i]] <-  field_foams_concentrations_df[[i]] * (field_foams_concentrations_df$drift_multiplier)^-1
  }
}
```

## CCB Correction

Now that drift has been accounted for, the next, much simpler step is to subtract out the blank concentrations as determined from the CCB samples. The mean of each analyte in the CCB samples is subtracted from the drift-corrected values, which is stored in the dataframe `drift_corrections`.

```{r}
# Create a separate dataframe with just CCB data
CCB_data <-  subset(drift_corrections, id == "CCB")

# Averaging the CCB numbers
background_corrections <- data_frame(n = drift_corrections$n, id = drift_corrections$id)
for (i in names(CCB_data)) {
  if (is.numeric(CCB_data[[i]]) && i != "n") {
    mean_i <- mean(CCB_data[[i]])
    background_corrections[[i]] = drift_corrections[[i]] - mean_i
  }
}
```

Next, `corrected_data` is created from `background_corrections` that with sample ID information followed by columns ordered alphabetically by analyte. The names of the analytes are stored in the vector `analytes`.

```{r}
# Remove Y column
ids <-  background_corrections$id
corrected_data <- subset(background_corrections, select = -c(3))
id_info <- subset(background_corrections, select = c(1,2))
corrected_data <- corrected_data[,3:ncol(corrected_data)]
corrected_data <- corrected_data[,order(colnames(corrected_data))]
analytes <- colnames(corrected_data)
corrected_data <- cbind(id_info, corrected_data)
```


## QA/QC Testing

There are several quality assurance and quality control checks for the run as a whole. Several are contingent on `IDL`, `MDL`, and `PQL_LOQ_MRL`, which are loaded in from saved data. Each run may only contain a subset of the full host analytes contained in these three vectors, so each is pared down to only contain the desired analytes.

```{r}
# Make sure that only analytes present are used in the analysis
IDL <- IDL[analytes]
MDL <- MDL[analytes]
PQL_LOQ_MRL <- PQL_LOQ_MRL[analytes]
```

`quality()` is a custom function made for the purpose of comparing a given control sample to a range in which its analytes should fall into. 

The `vec` argument corresponds to the analyte concentrations of a given sample/control. `comp` is ether a single desired concentration for all analytes to be at, or a vector of such concentrations with one concentration for each analyte. `lower` and `upper` represent lower and upper boundaries for acceptable concentrations. For example, if you wanted a control to be within 15% of 100, you would have `quality(x, comp = 100, lower = 0.85, upper = 1.15)`. The function returns a value of "Valid" or "Invalid" for each analyte of a given vector based on whether or not it falls in the desired range.

```{r}
# General function that can compare several quanitities to its desired range
# Ex. vec = c(83, 84, 85, 86), comp = 100, lower = 0.85, upper = 1.15
# Invalid Invalid Valid Valid
quality <-  function(vec, comp, lower, upper){
  validity <- c()
  lower_limit <- comp * lower
  upper_limit <- comp * upper
  if (length(comp) > 1){
    for (i in 1:length(vec)){
      if (vec[i] >= lower_limit[i] && vec[i] <= upper_limit[i]){
        validity <- c(validity, "Valid")
      } else{
        validity <- c(validity, "Invalid")
      }
    }
  } else{
    for (n in vec){
      if (n >= lower_limit && n <= upper_limit){
        validity <- c(validity, "Valid")
      } else{
        validity <- c(validity, "Invalid")
      }
    }
  }
  return(validity)
}
```

`data_range` is defined as the range of columns in `corrected_data` that actually contains analyte data.

```{r}
# Column number where actual concentration data begins
data_range <- 3:length(corrected_data)
```

`QA_QC` is a list-type object that will store information about the various checks. 

The IPC-500, LFM, and CCB checks all look at multiple samples. Their corresponding variables keep track of how many samples of each have gone through the check.

```{r}
QA_QC <- list()
# Keeps track of how many of each samples have been checked
IPC500 <- 1
LFM <- 1
CCB <- 1
```

For the IPC-500, QCS, SIC, and LFB checks, iron and zinc are spiked to be at higher concentrations than other analytes. Therefore, the corresponding samples need to compared to a vector with different values for iron and zinc, whereas other tests may only need to be compared to a number.

```{r}
IPC_comp <- c()
QCS_comp <- c()
SIC_comp <- c()
LFB_comp <- c()
for (i in colnames(corrected_data[,data_range])) {
  if (startsWith(i, "Fe")) {
    IPC_comp <- c(IPC_comp, 2500)
    QCS_comp <- c(QCS_comp, 1250)
    SIC_comp <- c(SIC_comp, 125 *5)
    LFB_comp <- c(LFB_comp, 125 *5)
    
    
  } else if (startsWith(i, "Zn")){
    IPC_comp <- c(IPC_comp, 1250)
    QCS_comp <- c(QCS_comp, 625)
    SIC_comp <- c(SIC_comp, 125 * 2.5)
    LFB_comp <- c(LFB_comp, 125 *2.5)
    
  } else {
    IPC_comp <- c(IPC_comp, 500)
    QCS_comp <- c(QCS_comp, 250)
    SIC_comp <- c(SIC_comp, 125)
    LFB_comp <- c(LFB_comp, 125)
  }
}
```

Below is the actual QA/QC process using a for loop and conditional statements. The results are stored in `QA_QC`.

Further information about each test is found commented into the code.

```{r}
# Actual QA/QC procress
for (i in 1:nrow(corrected_data)){
  id <- corrected_data$id[i]
  # IPC-500 (1st one) must be 95-105% of 500 ppb of analyte
  if (id == "IPC-500" && IPC500 == 1) {
    comp <- 
      QA_QC[["IPC1"]] <- quality(vec = corrected_data[i, data_range], comp = IPC_comp, lower = 0.95, upper = 1.05)
    IPC500 <- IPC500 + 1
    # IPC-500 (2nd, 3rd, 4th, etc.) must be 90-110% of 500 ppb of analyte
  } else if (id == "IPC-500") {
    name <- paste("IPC", IPC500, sep = "")
    QA_QC[[name]] <- quality(vec = corrected_data[i, data_range], 
                             comp = IPC_comp, 
                             lower = 0.90, 
                             upper = 1.10)
    IPC500 <- IPC500 + 1
    # LFM(-1,-2,-3) must be 85-115% of predicted spike concentration
    # sample id must contain LFM
  } else if (grepl("LFM", id)){
    name <- paste("LFM", LFM, sep = "")
    comp <- corrected_data[i-1, data_range]/2 + 125
    comp["Zn206.200(µg/L)"] <- comp["Zn206.200(µg/L)"] - 125 + (125 * 2.5)
    comp["Fe238.204(µg/L)"] <- comp["Fe238.204(µg/L)"] - 125 + (125 * 5)
    QA_QC[[name]] <- quality(vec = corrected_data[i, data_range], 
                             comp = comp, 
                             lower = 0.85, 
                             upper = 1.15)
    LFM <- LFM + 1
    # CCB must be < IDL
  } else if(grepl("CCB", id)){
    name <- paste("CCB", CCB, sep = "")
    result <- as.character(corrected_data[i, data_range] < IDL)
    result <- gsub("FALSE", "Invalid", x = result)
    result <- gsub("TRUE", "Valid", x = result)
    QA_QC[[name]] <- result
    CCB <- CCB + 1
    # LRB must be < 2.2 xMDL
  } else if(grepl("LRB", id)){
    result <- as.character(corrected_data[i, data_range] < (2.2*MDL))
    result <- gsub("FALSE", "Invalid", x = result)
    result <- gsub("TRUE", "Valid", x = result)
    QA_QC[["LRB"]] <- result
    # SIC-125 must be 80-120% of 125 ppb of analyte.
  } else if (grepl("SIC", id)) {
    QA_QC[["SIC"]] <- quality(vec = corrected_data[i, data_range], 
                              comp = SIC_comp, 
                              lower = 0.80, 
                              upper = 1.20)
    # QCS-250 must be 95-105% of 250 ppb of analyte
  } else if (grepl("QCS", id)) {
    QA_QC[["QCS"]] <- quality(vec = corrected_data[i, data_range], 
                              comp = QCS_comp, 
                              lower = 0.95, 
                              upper = 1.05)
    # LFB must be 85-115% of 125 ppb of analyte
  } else if (grepl("LFB", id)) {
    QA_QC[["LFB"]] <- quality(vec = corrected_data[i, data_range], 
                              comp = LFB_comp, 
                              lower = 0.85, 
                              upper = 1.15)
  }
}
```

`QA_QC` is then converted into a dataframe to easily convert it into an Excel sheet later on. But before the dataframe is finalized, the three tests with multiple samples have to be removed from the rest of test results.

```{r}
# Finalizing the QA/QC dataframe
QA_QC_df <- as.data.frame(t(as.data.frame(QA_QC)))
colnames(QA_QC_df) <- colnames(corrected_data[data_range])
QA_QC_df$id <- rownames(QA_QC_df)
IPCQ <- subset(QA_QC_df, startsWith(id, "IPC"), select = -c(id))
LFMQ <- subset(QA_QC_df, startsWith(id, "LFM"), select = -c(id))
CCBQ <- subset(QA_QC_df, startsWith(id, "CCB"), select = -c(id))
QA_QC_final <- subset(QA_QC_df, !startsWith(id, "IPC") & !startsWith(id, "LFM") & !startsWith(id, "CCB"), select = -c(id))
```

These multiple sample runs are combined such that if for any individual sample an analyte was determined to be invalid, for that test, that analyte is invalid. All samples must say a particular analyte is valid for it to be considered as such.

The final QA/QC results are stored in the dataframe `QA_QC_final`.

```{r}
# Combine IPC, LFM, and CCB runs
IPC_tot <- c()
for (i in names(IPCQ)){
  if (length(unique(IPCQ[[i]])) == 1 && unique(IPCQ[[i]]) == "Valid"){ 
    IPC_tot <- c(IPC_tot, "Valid")
  } else {
    IPC_tot <- c(IPC_tot, "Invalid")
  }
}

LFM_tot <- c()
for (i in names(LFMQ)){
  if (length(unique(LFMQ[[i]])) == 1 && unique(LFMQ[[i]]) == "Valid"){
    LFM_tot <- c(LFM_tot, "Valid")
  } else {
    LFM_tot <- c(LFM_tot, "Invalid")
  }
}
CCB_tot <- c()
for (i in names(CCBQ)){
  if (length(unique(CCBQ[[i]])) == 1 && unique(CCBQ[[i]]) == "Valid"){
    CCB_tot <- c(CCB_tot, "Valid")
  } else {
    CCB_tot <- c(CCB_tot, "Invalid")
  }
}
# Final output for QA/QC
QA_QC_final <- rbind(QA_QC_final, IPC = IPC_tot, LFM = LFM_tot, CCB = CCB_tot)
```

## The Analysis

Time for the actual analysis. The goal is to see if the concentration of analyte found in water samples is significantly greater than the background foams themselves, accounting for the weight of foams. If so, the concentrations of these samples will reported.

First, we read in the weights of the desired foams from an Excel file and change the format of the sample name to match the data already in use. 

```{r}
weights_df <- read_excel(weights)
weights_df$`Foam ID` <- gsub(" ", "-", weights_df$`Foam ID`)
weights_df$`Foam ID` <- paste("A", weights_df$`Foam ID`, sep = "")
```

We need to pare down the drift and background corrected data to only the desired water samples, which is found in `sample_df`. Many sample weights are found in the weights dataframe, and so it is pared down in a similar manner into `sample_weights`. `sample_num` contains just the raw sample numbers (17 as opposed to A17-1C).

```{r}
corrected_data$id <- as.character(corrected_data$id)
sample_df <- subset(corrected_data, startsWith(id, sample_start) & endsWith(id, sample_end), select = -c(n))
sample_df <- sample_df[order(sample_df$id),]
sample_weights <- subset(weights_df, `Foam ID` %in% sample_df$id)
sample_num <- c()
for (name in sample_df$id){
  parts <- strsplit(name, "-")
  parts2 <- strsplit(parts[[1]][1], "A")
  sample_num <- c(sample_num, parts2[[1]][2])
}
```

The samples must then be divided by the weight of the "A" piece of foam from which they were taken. This weight is an average of the associated foams (mean weight of A17-1C, A17-2C, and A17-3C weights). 

The error is measured in terms of standard deviation.

```{r}
# Divide by analyzed foam weight
sample_weights$sample_num <- as.factor(sample_num)
Aweights_means <- with(sample_weights, tapply(`Weight of half of foam analyzed (A) (g)`, sample_num, mean))
Tweights_means <- with(sample_weights, tapply(`Total weight of foam (g)`, sample_num, mean))
Aweights_sds <- with(sample_weights, tapply(`Weight of half of foam analyzed (A) (g)`, sample_num, sd))
Tweights_sds <- with(sample_weights, tapply(`Total weight of foam (g)`, sample_num, sd))
sample_norm <- cbind(sample = as.numeric(sample_num),sample_df[,2:ncol(sample_df)]/sample_weights$`Weight of half of foam analyzed (A) (g)`)
sample_df <- cbind(sample = as.numeric(sample_num), sample_df[,2:ncol(sample_df)])
sample_list <- c(unique(sample_norm$sample))
```

Background foam weights and concentrations are pre-loaded into the R environment via `load("IPC-analysis-app/data/IPC_data.Rdata")` 

Different samples used different types of foams (1: samples 1-198; 2: samples 199-212; 3: samples 213-316), so the weights and concentrations for the three foam types must be separated into their own dataframes.

```{r}
# Divide background foam concentrations by foam weight
bg1_weights <- subset(background1, 
                      select = c(`Ahalfweight(g)`,
                                 `Bhalfweight(g)`,
                                 `Totalweight(g)`)
)
bg1_conc <- subset(background1, select = analytes)
bg2_weights <- subset(background2, 
                      select = c(`Ahalfweight(g)`,
                                 `Bhalfweight(g)`,
                                 `Totalweight(g)`)
)
bg2_conc <- subset(background2, select = analytes)
bg3_weights <- subset(background3, 
                      select = c(`Ahalfweight(g)`,
                                 `Bhalfweight(g)`,
                                 `Totalweight(g)`)
)
bg3_conc <- subset(background3, select = analytes)
```

The same weight-normalization we did for the samples must now happen for background analyte concentrations and their associated weights.

```{r}
# Grab concentrations, means and SDs for the background data
bg1_conc <- bg1_conc[,order(colnames(bg1_conc))]
bg1_conc_mean <- colMeans(bg1_conc, na.rm = T)
bg1_conc_sd <- apply(bg1_conc, 2, sd, na.rm=TRUE)
bg2_conc <- bg2_conc[,order(colnames(bg2_conc))]
bg2_conc_mean <- colMeans(bg2_conc, na.rm = T)
bg2_conc_sd <- apply(bg2_conc, 2, sd, na.rm=TRUE)
bg3_conc <- bg1_conc[,order(colnames(bg3_conc))]
bg3_conc_mean <- colMeans(bg3_conc, na.rm = T)
bg3_conc_sd <- apply(bg3_conc, 2, sd, na.rm=TRUE)
bg1_norm <- bg1_conc/bg1_weights$`Bhalfweight(g)`
bg2_norm <- bg2_conc/bg2_weights$`Bhalfweight(g)`
bg3_norm <- bg3_conc/bg3_weights$`Bhalfweight(g)`
```

For any given sample, each analyte concentration mean has to compared to background foam concentration mean in an independent samples 1-tailed t-test, where we are testing the hypothesis that the mean concentration in the sample is greater than the mean concentration in the background foam. The custom function `sig_test()` lets you run all of the necessary t-tests for each analyte and returns each p-value in a vector. If a sample only has one replicate, a t-test will not work, and a p-value of 1 will be returned. 

Its input `sample_df` is a subset of the weight-normalized sample concentrations dataframe only containing one sample, and `background_df` is the corresponding weight-normalized background concentrations.

```{r}
# Defines a function to compare values
sig_test <- function(sample_df, background_df) {
  p_values <- c()
  for (i in colnames(sample_df)){
    if (length(sample_df[[i]]) < 2){
      p_values <- c(p_values, 1)
    } else {
      p_values <- c(p_values, t.test(sample_df[[i]], background_df[[i]], alternative = "greater")$p.value)
    }
  }
  return(p_values)
}
```

This next custom function `final_label()` decides how to display each analyte in the final spreadsheet. An analyte can one of the following categories: "NotSignificant" (not significantly greater than the background foam concentration, alpha = 0.05), "Non-detect" (concentration less than or equal to MDL), "Present" (greater than MDL but less than or equal to PQL,LOQ,MRL), and actually displaying the concentration (greater than PQL,LOQ,MRL).

The inputs to this function are simply vectors containing booleans (TRUE or FALSE) stating whether or not the stated label applies to a particular analyte, and the final concentrations (`final_conc`) for a given sample.

```{r}
# How to label each analyte of each sample for the final main spreadsheet
final_label <- function(signif_v, nondetect, only_detect, give_value, final_conc) {
  final <- c()
  for (i in 1:length(signif_v)) {
    if (!signif_v[i]) {
      final <- c(final, "NotSignificant")
    } else if (nondetect[i]) {
      final <- c(final, "Non-detect")
    } else if (only_detect[i]) {
      final <- c(final, "Present")
    } else if (give_value[i]){
      final <- c(final, final_conc[i])
    }
  }
  return(final)
}
```

Here's where those functions get put to use: in a monster for-loop. Subsets are made only containing a single sample from the `sample_norm` (weight-normalized) and `sample-df` (non-normalized concentrations) dataframes. The non-normalized concentrations (`conc_mean`) are used to determine the labeling discussed above via boolean vectors, besides the significance vector.

The background foam values compared to depend on the sample number, as described above. The non-normalized background concentration mean is then subtracted from the sample mean. This can be negative if sample concentration mean is lower than the background concentration. The ratio of A side of the foam's weight to the total foam weight is then incorporated to find the final concentration. The error propagation takes into account each of these. 

Matrices are then constructed to build toward the sheets in the final Excel spreadsheet. Matrices are easier at this point to deal with than dataframes.

```{r}
for (i in 1:length(sample_list)) {
  sample_number <- sample_list[i]
  norm_df <- paste(sample_start, sample_number, "_norm" ,sep = "")
  conc_df <- paste(sample_start, sample_number, "_conc" ,sep = "")
  assign(norm_df, subset(sample_norm, sample == sample_number, select = -c(sample)))
  assign(conc_df, subset(sample_df, sample == sample_number, select = -c(sample)))
  conc_mean <- apply(get(conc_df), 2, mean)
  conc_sd <- apply(get(conc_df), 2, sd)
  # for labels
  nondetect <- conc_mean <= MDL
  only_detect <- (conc_mean > MDL) & (conc_mean <= PQL_LOQ_MRL)
  give_value <- conc_mean > PQL_LOQ_MRL
  
  if (sample_number < 198) {
    p_values <- sig_test(sample_df = get(norm_df), background_df = bg1_norm)
    final_conc <- (conc_mean - bg1_conc_mean)
    final_sd <- sqrt(conc_sd^2 + bg1_conc_sd^2 - 2 * conc_sd * bg1_conc_sd)
    
  } else if (sample_number >= 198 & sample_number < 213) {
    p_values <- sig_test(sample_df = get(norm_df), background_df = bg2_norm)
    final_conc <- (conc_mean - bg2_conc_mean)
    final_sd <- sqrt(conc_sd^2 + bg2_conc_sd^2 - 2 * conc_sd * bg2_conc_sd)
    
  } else if (sample_number >= 213 & sample_number < 317){
    p_values <- sig_test(sample_df = get(norm_df), background_df = bg3_norm)
    final_conc <- (conc_mean - bg3_conc_mean)
    final_sd <- sqrt(conc_sd^2 + bg3_conc_sd^2 - 2 * conc_sd * bg3_conc_sd)
  }
  weight_ratio <- 3/7 * (Tweights_means[as.character(sample_number)]/Aweights_means[as.character(sample_number)])
  weight_sd <-  3/7 * sqrt((Tweights_sds[as.character(sample_number)]/Tweights_means[as.character(sample_number)])^2 + (Aweights_sds[as.character(sample_number)]/Aweights_means[as.character(sample_number)])^2 - 2 * (Aweights_sds[as.character(sample_number)] * Tweights_sds[as.character(sample_number)])/(Aweights_means[as.character(sample_number)] * Tweights_means[as.character(sample_number)]))
  final_sd <- sqrt((final_sd/final_conc)^2 + (weight_sd/weight_ratio)^2 + 2 * (weight_sd * final_sd)/(weight_ratio * final_conc))
  final_conc <- final_conc * weight_ratio
  signif_v <- p_values < 0.05
  final_important <- final_label(signif_v, nondetect, only_detect, give_value, final_conc)
  # Have to initialize matrices the first time through the loop
  if (i == 1){
    all_p_values <- matrix(p_values, ncol = length(p_values))
    signif_m <- matrix(signif_v, ncol = length(p_values))
    nondetect_m <- matrix(nondetect, ncol = length(nondetect))
    only_detect_m <- matrix(only_detect, ncol = length(only_detect))
    give_value_m <- matrix(give_value, ncol = length(give_value))
    final_conc_m <- matrix(final_conc, ncol = length(final_conc))
    final_sd_m <- matrix(final_sd, ncol = length(final_sd))
    final_important_m <- matrix(final_important, ncol = length(final_important))
  # Matrices are already initialized now on subsequent passes
  } else {
    all_p_values <- rbind(all_p_values, p_values)
    signif_m <- rbind(signif_m, signif_v)
    nondetect_m <- rbind(nondetect_m, nondetect)
    only_detect_m <- rbind(only_detect_m, only_detect)
    give_value_m <- rbind(give_value_m, give_value)
    final_conc_m <- rbind(final_conc_m, final_conc)
    final_important_m <- rbind(final_important_m, final_important)
    final_sd_m <- rbind(final_sd_m, final_sd)
  }
}
```

All of matrices can now be converted to dataframes. Column names have to be occasionally altered. 

At this point, the R Shiny version of the script ends, and these data frames are converted into an Excel file.

```{r}
# convert many matrices to data frames; makes it easier to export as an excel file 
all_p_values <- as.data.frame(all_p_values, row.names = paste(sample_start,as.character(sample_list), sep = ""))
colnames(all_p_values) <- colnames(bg1_norm)
signif_df <- as.data.frame(signif_m, row.names = paste(sample_start,as.character(sample_list), sep = ""))
colnames(signif_df) <- colnames(bg1_norm)
nondetect_df <- as.data.frame(nondetect_m, row.names = paste(sample_start,as.character(sample_list), sep = ""))
only_detect_df <- as.data.frame(only_detect_m, row.names = paste(sample_start,as.character(sample_list), sep = ""))
give_value_df <- as.data.frame(give_value_m, row.names = paste(sample_start,as.character(sample_list), sep = ""))
final_conc_df <- as.data.frame(final_conc_m, row.names = paste(sample_start,as.character(sample_list), sep = ""))
final_important_df <- as.data.frame(final_important_m, row.names = paste(sample_start,as.character(sample_list), sep = ""))
colnames(final_important_df) <- colnames(bg1_norm)
final_sd_df <- as.data.frame(final_sd_m, row.names = paste(sample_start,as.character(sample_list), sep = ""))
drift_corrections_temp <- subset(drift_corrections, select = -c(n, id))
drift_corrections_temp <- drift_corrections_temp[,order(colnames(drift_corrections_temp))]
drift_corrections_df <- cbind(id = ids, drift_corrections_temp)
corrected_data_df <- subset(corrected_data, select = -c(n))
```

However, in the script version, the code to write an excel file is included.

```{r}
write.xlsx(final_important_df, file = output, sheetName = "FinalAnalysis", append = FALSE, col.names = T, row.names = T)
write.xlsx(QA_QC_final, file = output, sheetName = "QA_QC", append = TRUE, col.names = T, row.names = T)
write.xlsx(drift_corrections_df, file = output, sheetName = "DriftCorrection", append = TRUE, col.names = T, row.names = F)
write.xlsx(corrected_data_df, file = output, sheetName = "Drift_CCBCorrection", append = TRUE, col.names = T, row.names = F)
write.xlsx(all_p_values, file = output, sheetName = "PvaluesComparingToBackground", append = TRUE, col.names = T, row.names = T)
write.xlsx(final_conc_df, file = output, sheetName = "AllFieldConcentrations", append = TRUE, col.names = T, row.names = T)
write.xlsx(final_sd_df, file = output, sheetName = "AllFieldSDs", append = TRUE, col.names = T, row.names = T)
```

